---
title: "Databases"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Load/install packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, DBI, dbplyr, RSQLite, bigrquery, hrbrthemes, nycflights13, glue)
## My preferred ggplot2 theme (optional)
theme_set(hrbrthemes::theme_ipsum())
```

#### Learning and practicing SQL

**dplyr**'s `show_query()` function is a great way to get started. So is the [SQL translation vignette](https://dbplyr.tidyverse.org/articles/sql-translation.html).

The best way to learn is to write your own queries. The [BigQuery UI](https://console.cloud.google.com/bigquery) is helpful for this.

#### Resources

* ["Getting Global Fishing Watch Data from Google BigQuery using R](http://jsmayorga.com/post/getting-global-fishing-watch-from-google-bigquery-using-r)
* Julia Evans' [*Become A Select Star*](https://wizardzines.com/zines/sql/)
* Official BigQuery [documentation](https://cloud.google.com/bigquery/docs/) overviews the functions and syntax for "standard SQL"

#### Contents

- [Databases and the tidyverse](#databases-and-the-tidyverse)

- [Getting started with SQLite](#getting-started-with-sqlite)

  * [Connecting to a database](#connecting-to-a-database)
  * [Generating queries](#generating-queries)
  * [Laziness as a virtue](#laziness-as-a-virtue)
  * [Collect the data into your local R environment](#collect-the-data-into-your-local-r-environment)
  
- [Using SQL directly in R](#using-sql-directly-in-r)

  * [Translate with dplyr::show_query()](#translate-with-dplyr--show-query--)
  * [Option 1: Use R Markdown `sql` chunks](#option-1--use-r-markdown--sql--chunks)
  * [Option 2: Use DBI::dbGetQuery()](#option-2--use-dbi--dbgetquery--)
  * [Recommendation: Use glue::glue_sql()](#recommendation--use-glue--glue-sql--)
  * [Disconnect](#disconnect)
  
- [Google BigQuery](#google-bigquery)

  * [Example 1: US birth data](#example-1--us-birth-data)
  * [Example 2: Global Fishing Watch](#example-2--global-fishing-watch)


Many "big data" problems could be described as "small data problems in disguise", meaning the data we care about is just a small subset or aggregation of a larger dataset. For example, we might want to access US Census data...but only for a few counties along a state border. Or, we might want to analyze climate data collected from many weather stations...but aggregated to the national/monthly level. In these cases, the bottleneck is interacting with the original data, which is too large to fit in memory. We can store and access this data with **relational databases**.

Databases can exist locally or remotely (though remotely is more common). They are stored *on-disk* somewhere, rather than in-memory. We extract the information we want by submitting a *query* to the database. In the query, we can request that the data be manipulated or subsetted in a certain way before it is delivered to us.

**A table in a database is like a data frame in an R list.**

## Databases and the tidyverse

**dbplyr** allows for direct communication with databases from your local R environment. It's provides a database backend to **dplyr**. It depends on the **DBI** package, which provides a common interface to allow **dplyr** to work with many different databases using the same type of code.

For **DBI** to work, you must install a specific backend package for the type of database you want to connect to. [Here](https://db.rstudio.com/dplyr/#getting-started) is a list of commonly used backend packages. We will use:

* **RSQLite**, which embeds a SQLite database.
* **bigrquery**, which connects to Google BigQuery.

SQLite is a lightweight SQL database engine that can exist on our local computers (no need to connect to a server). BigQuery is a convenient way to get data if you already have a Google Cloud account.

## Getting started with SQLite

(For more, see RStudio's [Databases using dplyr](https://db.rstudio.com/dplyr) tutorial.)

### Connecting to a database

Start by opening an (empty) database connection via the `DBI::dbConnect()` function, which we'll call `con`. Note, we are calling the **RSQLite** package in the backgroud for the SQLite backend and telling R this is a local connection that exists in memory.

```{r}
# library(DBI)
con <- dbConnect(RSQLite::SQLite(), path = ":memory:")
```

The first argument is the database backend, i.e. `RSQLite::SQLite()` since that's what we're using in this case. SQLite only needs one other argument: the `path` to the database. We use the special string `":memory:"` which causes SQLite to create a temporary, in-memory database.

Our makeshift database is empty, so let's copy in the *flights* dataset from the **nycflights13** package. We'll use `dplyr::copy_to()` to do that.

The indexes enable efficient database performance. These are set by the database host platform or maintainer in normal applications.

```{r}
# library(nycflights13)
# library(dplyr)

copy_to(
  dest = con,
  df = nycflights13::flights,
  name = "flights",
  temporary = FALSE,
  indexes = list(
    c("year", "month", "day"),
    "carrier",
    "tailnum",
    "dest"
    )
  )
```

Now that we've copied the data to the database, we can reference it via `dplyr::tbl()`:

```{r}
# library(dbplyr)

flights_db <- tbl(con, "flights")
flights_db
```

### Generating queries

**dplyr** auto-translates tidyverse style code into SQL. Some examples:

```{r}
## Select some columns
flights_db %>% select(year:day, dep_delay, arr_delay)
```

```{r}
## Filter according to a condition
flights_db %>% filter(dep_delay > 240)
```

```{r}
## Get the mean delay by destination
flights_db %>%
  group_by(dest) %>%
  summarise(delay = mean(dep_time)) %>%
  arrange(desc(delay))
```

### Laziness as a virtue

**dplyr** tries to be as "lazy" as possible. Your R code is translated into SQL and executed in the database, not in R. This is good because:

* It never pulls data into R unless you explicitly ask.
* It delays doing any work until the last possible moment: it collects everything you want to do and sends it to the database in one step.

For example, if we're interested in the mean departure and arrival delays for each plane (unique tail number):

```{r}
tailnum_delay_db <- 
  flights_db %>%
  group_by(tailnum) %>%
  summarise(
    mean_dep_delay = mean(dep_delay),
    mean_arr_delay = mean(arr_delay),
    n = n()                            ## Count number of obs.
  ) %>%
  arrange(desc(mean_arr_delay)) %>%
  filter(n >= 100)
```

This sequence of operations never even touches the database. It's not until you ask for the data (e.g. by calling `tailnum_delay_db`) that **dplyr** actually generates the SQL and requests the results from the database. Even then, it limits the number of rows it pulls:

```{r}
tailnum_delay_db
```

### Collect the data into your local R environment

Typically you'll iterate a few times before figuring out what data you need from the database. Then, use `collect()` to pull it into a local data frame:

```{r}
tailnum_delay <- 
  tailnum_delay_db %>%
  collect()
tailnum_delay
```
Now we see it has 1,218 rows. It's stored as a data frame, so we can use it just like any other.

Question: What is the relationship between late departures and late arrivals?

```{r}
tailnum_delay %>%
  ggplot(aes(x = mean_dep_delay,
             y = mean_arr_delay,
             size = n)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, col = "orange") +
  coord_fixed()
```

Assuming we're finished querying our SQLite database, we'd disconnect from it by calling `DBI::dbDisconnect(con)`.

## Using SQL directly in R

### Translate with dplyr::show_query()

Behind the scenes, **dplyr** is translating your R code to SQL. Use `show_query()` to display the SQL code that was used to generate a queried table:

```{r}
tailnum_delay_db %>%
  show_query()
```

There are some translation artifacts, because dplyr implements safeguards to ensure the translated SQL will absolutely work. This explains the repeated `SELECT` commands.

SQL uses a *lexical* order of operations that doesn't always preserve the *logical* order of operations, meaning the way we write a query is not the same way we think through a query. (SQL has a strict "order of execution", so there is a hierarchy of commands.) [Here](https://wizardzines.com/zines/sql/samples/from.png) is a graphic explaining it, from the zine ["Become a Select Star"](https://wizardzines.com/zines/sql/).

The main confusion point is, we always write `SELECT` first, even though `SELECT` doesn't come into play until later in the query's logic.

The **DBI** package lets you write and submit SQL queries directly from R.

```{r}
## Show the equivalent SQL query for a group of dplyr commands
flights_db %>% filter(dep_delay > 240) %>% head(5) %>% show_query()
```

Note: the backticks around object names, and the parentheses around `WHERE`, are just safeguards that are not always necessary.

### Option 1: Use R Markdown `sql` chunks

Just use the header ```` ```{sql, connection = con} ```` instead of ```` ```{r} ````. (Change `con` to your own database connection.)

Once knitted, this looks like:

```{sql, connection = con}
SELECT *
FROM flights
WHERE dep_delay > 240
LIMIT 5
```

### Option 2: Use DBI::dbGetQuery()

To run SQL queries in regular R scripts, use `DBI::dbGetQuery()`:

```{r}
## Run the query using SQL directly on the connection.
dbGetQuery(con, "SELECT * FROM flights WHERE dep_delay > 240 LIMIT 5")
```

### Recommendation: Use glue::glue_sql()

Grant recommends the `glue_sql()` function from **glue**, part of the tidyverse. It provides a more integrated approach that lets you use local R variables in your SQL queries, and divide long queries into sub-queries.

```{r}
## Using R variables in a SQL query

# library(glue)

## Some local R variables
tbl <- "flights"
d_var <- "dep_delay"
d_thresh <- 240

## The "glued" SQL query string
sql_query <-
  glue_sql("
  SELECT *
  FROM {`tbl`}
  WHERE ({`d_var`} > {d_thresh})
  LIMIT 5
  ", 
  .con = con
  )

## Run the query
dbGetQuery(con, sql_query)
```
`glue_sql()` really pays off when working with large, nested queries. [Documentation here](https://glue.tidyverse.org/reference/glue_sql.html).

### Disconnect

```{r}
## Terminate connection with the database
dbDisconnect(con)
```

## Google BigQuery

There are tons of public datasets available on BigQuery. The console has a nice web UI too.

Here is a [tutorial](https://towardsdatascience.com/bigquery-without-a-credit-card-discover-learn-and-share-199e08d4a064) for using BigQuery to determine who is the most popular Alan (measured by Wikipedia page views).

To use the **bigrquery** package, you must provide your GCP project billing ID. We stored our credentials in the `.Renviron` file in our home directory during the cloud computing lecture. In that case, you can call it using `Sys.getenv()`.

```{r}
library(bigrquery)
billing_id <- Sys.getenv("GCE_DEFAULT_PROJECT_ID")  ## Replace with your project ID if this doesn't work
```

### Example 1: US birth data

Establish a connection using `DBI::dbConnect()`:
```{r}
# library(DBI)
# library(dplyr)

bq_con <- 
  dbConnect(
    bigrquery::bigquery(),  ## Specify the BigQuery backend
    project = "publicdata",
    dataset = "samples",
    billing = billing_id  ## Provide credentials
    )
```

This connection works for any tables in the database. Set up auto-authentication, then list the tables:

```{r}
bq_auth(path = Sys.getenv("GCE_AUTH_FILE")) ## Or you could authenticate manually instead
 
dbListTables(bq_con)
```

Create a pointer to the "natality" table in our local environment:
```{r}
natality <- tbl(bq_con, "natality")
```

Summarize as mean, and collect data into our session:
```{r}
bw <- 
  natality %>%
  filter(!is.na(state)) %>% ## remove some outliers
  group_by(year) %>%
  summarise(weight_pounds = mean(weight_pounds, na.rm=T)) %>%
  collect()
```

```{r}
bw %>%
  ggplot(aes(year, weight_pounds)) +
  geom_line()
```

Query again, but grouping by year, state, and gender:
```{r}
bw_st <- 
  natality %>%
  filter(!is.na(state)) %>%
  group_by(year, state, is_male) %>%
  summarise(weight_pounds = mean(weight_pounds, na.rm=T)) %>%
  mutate(gender = ifelse(is_male, "Male", "Female")) %>%
  collect()
```

```{r}
## Select arbitrary states to highlight
states <- c("CA","DC","OR","TX","VT")
## Rearranging the data will help with the legend ordering
bw_st <- bw_st %>% arrange(gender, year)

## Plot it
bw_st %>%
  ggplot(aes(year, weight_pounds, group=state)) + 
  geom_line(col="grey75", lwd = 0.25) + 
  geom_line(
    data = bw_st %>% filter(state %in% states), 
    aes(col=fct_reorder2(state, year, weight_pounds)),
    lwd=0.75
    ) +
  facet_wrap(~gender) + 
  scale_color_brewer(palette = "Set1", name=element_blank()) +
  labs(
    title = "Mean birth weight, by US state over time",
    subtitle = "Selected states highlighted",
    x = NULL, y = "Pounds",
    caption = "Data sourced from Google BigQuery"
    ) + 
  theme_ipsum(grid=F)
```

Remember to disconnect.
```{r}
dbDisconnect(bq_con)
```

### Example 2: Global Fishing Watch

```{r}
gfw_con <- 
  dbConnect(
    bigrquery::bigquery(),
    project = "global-fishing-watch",
    dataset = "global_footprint_of_fisheries",
    billing = billing_id
    )
```

```{r}
dbListTables(gfw_con)
```

```{r}
effort <- tbl(gfw_con, "fishing_effort")
effort
```

Which country fishes the most?

```{r}
effort %>%
  group_by(flag) %>%
  summarise(total_fishing_hours = sum(fishing_hours, na.rm = T)) %>%
  arrange(desc(total_fishing_hours)) %>%
  collect()
```

#### Date partitioning

Many tables in BigQuery are *date partitioned*, i.e. ordered according to timestamps of when the data were ingested. This provides a more cost-effective way to query datasets. But, it necessitates a minor tweak to how we query or manipulate this GFW data by date.

In native SQL we reference a special `_PARTITIONTIME` pseudo-column in order to, say, filter by date. There is no exact **dplyr** translation for this pseudo-column. But, we can specify it as a SQL variable in our **dplyr** call using backticks:

```{r}
effort %>%
  filter(
    `_PARTITIONTIME` >= "2016-01-01 00:00:00",
    `_PARTITIONTIME` < "2017-01-01 00:00:00"
    ) %>%
  group_by(flag) %>%
  summarise(total_fishing_hours = sum(fishing_hours, na.rm=T)) %>%
  arrange(desc(total_fishing_hours)) %>%
  collect()
```

#### Global fishing effort in 2016

```{r}
## Define the desired bin resolution in degrees (longitude/latitude)
resolution <- 1

globe <-
  effort %>%
  filter(
    `_PARTITIONTIME` >= "2016-01-01 00:00:00",
    `_PARTITIONTIME` <= "2016-12-31 00:00:00"
    ) %>%
  filter(fishing_hours > 0) %>%
  mutate(
    lat_bin = lat_bin/100,  ## Normalize data (not super important)
    lon_bin = lon_bin/100
    ) %>%
  mutate(  ## Find center of the 1-degree square bins, and round observations to the bin centers
    lat_bin_center = floor(lat_bin/resolution)*resolution + 0.5*resolution,
    lon_bin_center = floor(lon_bin/resolution)*resolution + 0.5*resolution
    ) %>%
  group_by(lat_bin_center, lon_bin_center) %>%
  summarise(fishing_hours = sum(fishing_hours, na.rm=T)) %>%
  collect()
```


```{r}
globe %>% 
  filter(fishing_hours > 1) %>% 
  ggplot() +
  geom_tile(aes(x=lon_bin_center, y=lat_bin_center, fill=fishing_hours))+
  scale_fill_viridis_c(
    name = "Fishing hours (log scale)",
    trans = "log",
    breaks = scales::log_breaks(n = 5, base = 10),
    labels = scales::comma
    ) +
  labs(
    title = "Global fishing effort in 2016",
    subtitle = paste0("Effort binned at the ", resolution, "° level."),
    y = NULL, x = NULL,
    caption = "Data from Global Fishing Watch"
    ) +
  theme_ipsum(grid=F) +
  theme(axis.text=element_blank())
```

```{r}
dbDisconnect(gfw_con)
```

