---
title: "webscraping-CSS"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(janitor)
library(tidyverse)
library(rvest)
library(lubridate)
```

* Web content is rendered either 1) server-side or 2) client-side.
* To scrape content rendered **server-side**, we must know the relevant **CSS selectors**.
* Find the CSS selectors with a browser extension or by using "inspect element".
* Use the `rvest` package to read the HTML doc into R and parse the relevant nodes.
  + A typical workflow: `read_html(URL) %>% html_nodes(CSS_SELECTORS) %>% html_table(fill=TRUE)`.
  + You may need other functions depending on content type (e.g. `html_text()`)

* Web scraping requires lots of experimenting and cleaning.

**Contents:**

* [Be polite](#be-polite)
* [rvest (for server-side)](#rvest-(for-server-side))
* [Exercise](#exercise)

## Be polite

The [polite](https://github.com/dmi3kno/polite) package provides tools to help maintain web etiquette while scraping (checking for permission, not hammering the host website with requests, etc).

## rvest (for server-side)

rvest is designed to work with web pages that are built server-side and thus requires knowledge of the relevant CSS selectors.

CSS has two main features:

* *Properties:* the "how" of display rules (e.g. fonts, styles, colors, page width)
* *Selectors:* the "what" of display rules (identify which properties should be applied to which elements; e.g. top-line text headers are labeled ".h1"; subheaderes are ".h2")

```{r}
# library(rvest)
m100 <- read_html("https://en.wikipedia.org/wiki/Men%27s_100_metres_world_record_progression")
m100
```

This creates an XML document containing the whole webpage.

Let's try to isolate a table on this page using SelectorGadget/ScrapeMate browser extension:
```{r, error=TRUE}
m100 %>%
html_nodes("div+ .wikitable td , div+ .wikitable th") %>%
html_table(fill=TRUE)
```
Didn't work. Sometimes you have to use Inspect Element instead to get the right CSS selector. Let's try this one:

```{r}
pre_iaaf <-
  m100 %>%
  html_nodes("table.wikitable:nth-child(8)") %>%
  html_table(fill=TRUE)
  
class(pre_iaaf)  ## Check class of the output
```
The output from this is actually a list of lists... so let's convert it to a data frame with `dplyr::bind_rows()`.
```{r}
# pre_iaaf <- pre_iaaf[[1]] ## Would also work

# library(dplyr)
pre_iaaf <-
  pre_iaaf %>%
  bind_rows() %>%
  as_tibble()

# library(janitor)
pre_iaaf <- clean_names(pre_iaaf) ## Clean column names
pre_iaaf
```

As seen above, there is an issue: Westergen's second entry is recorded incorrectly - where his name should be, his time is there instead. Let's write a simple fix:
```{r, error=TRUE}
## See if possible to convert the 'athlete' column to numeric. If possible, replace the cell with the preceding value.

pre_iaaf <-
  pre_iaaf %>%
  mutate(athlete = ifelse(is.na(as.numeric(athlete)), athlete, lag(athlete)))
```

Finally, convert date column from chr format to date format:
```{r}
# library(lubridate)
pre_iaaf <-
  pre_iaaf %>%
  mutate(date = mdy(date))
pre_iaaf
```

## Exercise

Scrape all the tables, combine, and make a plot.

```{r}
## Get next table
iaaf_1 <-
  m100 %>%
  html_nodes("table.wikitable:nth-child(14)") %>%
  html_table(fill=TRUE) %>%
  bind_rows() %>%
  as_tibble() %>%
  clean_names()
  
## Fix blank athlete, clean dates
iaaf_1 <- 
  iaaf_1 %>%
  mutate(athlete = if_else(athlete=="", lag(athlete), athlete)) %>%
  mutate(date = mdy(date))

## Fix blank dates
iaaf_1 <- 
  iaaf_1 %>%
  mutate(date = if_else(is.na(date), lag(date), date))
```

```{r}
## Get final table
iaaf_2 <- 
  m100 %>%
  html_nodes("table.wikitable:nth-child(19)") %>%
  html_table(fill = TRUE) %>%
  bind_rows() %>%
  as_tibble() %>%
  clean_names()
  
## Clean dates
iaaf_2 <- 
  iaaf_2 %>%
  mutate(date = mdy(date))

## Don't bother fixing the other issues, at least we can graph times against date.
```

```{r}
## Combine data frames
wr100 <- 
  bind_rows(
    pre_iaaf %>% select(time, date) %>% mutate(era = "Pre-IAAF"),
    iaaf_1 %>% select(time, date) %>% mutate(era = "Pre-automatic"),
    iaaf_2 %>% select(time, date) %>% mutate(era = "Modern")
  )

## Plot
wr100 %>%
  ggplot() +
  geom_point(mapping = aes(x = date, y = time, color = era)) +
  labs(title = "Men's 100 meters world record progression")
```



