---
title: 'Functions in R: Advanced'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(memoise, tidyverse, here)

square <- 
  function(x = 1) {
    x_sq <- x^2 
    df <- tibble(value=x, value_squared=x_sq)
    return(df)
  }
```

[Debug mode](#debug-mode)

[Catching user errors](#catching-user-errors)

  * [Function-specific control flow](#function-specific-control-flow)
  * [base::tryCatch()](#base--trycatch())
  
    + [Wrap tryCatch() around an entire function](#wrap-trycatch()-around-an-entire-function)
    + [Use tryCatch() inside a function](#use-trycatch()-inside-a-function)
    
  * [purrr::safely() and family](#purrr--safely()-and-family)
  
[Caching (memoisation)](#caching-(memoisation))

  * [Caching across R sessions](#caching-across-r-sessions)
  * [Verbose output](#verbose-output)
  
[Further resources](#further-resources)

## Debug mode

[Download cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf)

`debugonce()` lets you enter debug mode the next time you run the function you specify. (It's a version of `debug()` that prevents "debug hell" caused by recursion.)

You may also be prompted by RStudio to debug when there is an error.

## Catching user errors

Take for example this bad input:

```{r, error=TRUE}
square("one")
```

There are three main approaches to make your code user-proof:

### Function-specific control flow

We use `ifelse` to produce a warning or error message if the input argument is invalid:

```{r, error = TRUE}
square_ifelse <- 
  function(x = 1) {
    if (is.numeric(x)) { ## Check that the input is valid for our function.
      x_sq <- x^2
      df <- tibble(value=x, value_squared=x_sq)
      return(df)
    } else { ## Return a warning message if not.
      message("You must provide a numeric input variable.")
    }
  }
square_ifelse("one")
```
We can get a similar result with less code, using `stop()`:
```{r, error=TRUE}
square_stop <- 
  function (x = 1) {
    if (!is.numeric(x)) stop("You must provide a numeric input variable.")
    x_sq <- x^2
    df <- tibble(value=x, value_squared=x_sq)
    return(df)
  }
square_stop("one")
```

### base::tryCatch()

A more general option is to use `base::tryCatch()` for handling errors and warnings. Two examples:

#### Wrap tryCatch() around an entire function

We invoke R's built-in "error" class, which is passed to another built-in function "message". This tells R to produce our message when it recognizes that (any) error has occured inside our function.
```{r, error=TRUE}
tryCatch(
  square("three"), 
  error = function(e) message("Something went wrong. Did you try to square a string instead of a number?")
  )
```

This works, but throws away all our input because of a single error. Take this for example:
```{r, error=TRUE}
tryCatch(
  square(c(1,2,"three")), 
  error = function(e) message("Something went wrong. Did you try to square a string instead of a number?")
  )
```
Though most of our inputs were valid, we get a blanket-statement error message. Ideally we would only get an error message for the third input variable...

#### Use tryCatch() inside a function


```{r}
square_trycatch <-
  function (x = 1) {
    x_sq <- tryCatch(x^2, error = function(e) NA_real_) ## tryCatch goes here now. Produce an NA value if we can't square the input.
    df <- tibble(value=x, value_squared=x_sq)
    return(df)
  }
square_trycatch(c(1,2,"three"))
```
It returned NA for "three" but why also for 1 and 2?
```{r}
str(c(1,2,"three")) ## Return the structure of the data
```
R coerces every element in this vector to character strings, so each element got evaluated as NA.

Solution: use an input array that allows for different element types (i.e. a *list*.)

```{r}
map(list(1,2,"three"), square_trycatch)
```

We might want to bind the results into a single data frame, like with `purrr:map_df()`, but this produces errors because all the columns have to contain only one data type.

Solution: Coerce to numeric the offending input, within the function itself. This will introduce coercion warnings, but at least it won't fail:

```{r}
square_trycatch2 <- 
  function (x = 1) {
    x_sq <- tryCatch(x^2, error = function(e) NA_real_)
    df <- tibble(value=as.numeric(x), value_squared=x_sq) ## Convert input to numeric
    return(df)
  }
map_df(list(1,2,"three"), square_trycatch2)
```

### purrr::safely() and family

The tidyverse equivalent of `tryCatch()`. Related functions include `purrr::possibly()`.

```{r}
square_simple <-
  function (x = 1) {
    x_sq <- x^2
  }
square_safely <- safely(square_simple)
square_safely("three")
```
You can specify default behavior in case of an error:
```{r}
square_safely <- safely(square_simple, otherwise = NA_real_)
square_safely("three")
```

## Caching (memoisation)

Caching is good beacuse problems happen: crashes, malfunctions, power outages, memory limits, timeouts, invalid arguments buried in iterated input, etc.

The problems are worse with parallel computations: the whole program could run and then only reveal an error at the very end.

We use the [memoise](https://github.com/r-lib/memoise) package. (Memoisation is a particular form of caching that saves the results of expensive function calls, so we don't have to repeat them later.)

To see the benefits of caching, let's create a slow function and test without, and then with, caching:
```{r}
slow_square <- 
  function(x) {
    Sys.sleep(1)
    square(x)
  }
```
To enable caching, we feed the function to `memoise::memoise()`:
```{r}
library(memoise)
mem_square <- memoise(slow_square)
```

```{r}
## Run first time
system.time(
  m1 <- map_df(1:10, mem_square)
)
```
```{r}
## Run second time
system.time(
  m2 <- map_df(1:10, mem_square)
)
```
You could verify that the results are equal:
```{r}
all.equal(m1, m2)
```

If we run the function with some additional inputs, only the new inputs will add to the compute time:

```{r}
system.time(
  m3 <- map_df(1:15, mem_square)
)
```

### Caching across R sessions

Specify a dedicated cache directory for complex or time-consuming analyses that you want to be able to access across R sessions.

To enable caching that persists across sessions, specify a cache directory with the argument `cache = cache_filesystem(PATH)`. Recommend you use use a `.rcache/` naming pattern to keep your paths organized.
```{r}
## Cache directory path (which I've already created)
cache_dir <- here("R-funcs-adv/.rcache")

## (Re-)memoise our function with the persistent cache location
mem_square_persistent <-
  memoise(slow_square, cache = cache_filesystem(cache_dir))
```

### Verbose output

It's often helpful to add verbose prompts to memoised functions. The code below will:

1. Check for and load previously cached results. Print results to screen.
2. Run the memoised function on any inputs that haven't yet been evaluated. (These results will be cached.) Print results to screen.

```{r, eval=FALSE}
mem_square_verbose <- 
  function(x) {
    ## 1. Load cached data if already generated
    if (has_cache(mem_square_persistent)(x)) {
      cat("Loading cached data for x =", x, "\n")
      my_data <- mem_square_persistent(x)
      return(my_data)
    }
    
    ## 2. Generate new data if cache not available
    cat("Generating data from scratch for x =", x, "...")
    my_data <- mem_square_persistent(x)
    cat("ok\n")
    
    return(my_data)
  }
```

```{r, eval=FALSE}
system.time(
  m5 <- map_df(1:10, mem_square_verbose)
)
```

```
## Loading cached data for x = 1 
## Loading cached data for x = 2 
## Loading cached data for x = 3 
## Loading cached data for x = 4 
## Loading cached data for x = 5 
## Loading cached data for x = 6 
## Loading cached data for x = 7 
## Generating data from scratch for x = 8 ...ok
## Generating data from scratch for x = 9 ...ok
## Generating data from scratch for x = 10 ...ok
```

```
##    user  system elapsed 
##   0.022   0.003   6.031
```

## Further resources

* [Debugging techniques in RStudio](https://rstudio.com/resources/rstudioconf-2018/debugging-techniques-in-rstudio/) (a recorded talk, from RStudio)
* [Debugging with RStudio](https://support.rstudio.com/hc/en-us/articles/205612627-Debugging-with-RStudio) (also from RStudio)
* [Object of type 'closure' is not subsettable](https://rstudio.com/resources/rstudioconf-2020/object-of-type-closure-is-not-subsettable), Grant's favorite debugging talk
* [Debugging](https://adv-r.hadley.nz/debugging.html) chapter of *Advanced R*








